---
title: "Survival analysis: Coursework"
author: "Petteri Mäntymaa"
date: "May 3, 2019"
header-includes:
   - \usepackage{mathspec}
   - \usepackage{tikz}
output:
    pdf_document:
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse");library(ggplot2);library(survival);library(broom);library(eha); library(tinytex); library(survminer);library(KMsurv)
options("scipen"=0, "digits"=7)
set.seed(42)
```

#### Q1

#### a)
*Show that **the hazard function** is uniquely determined by the **mean residual time $r(t)$**.*

Mean residual time $r(t)$ at time $t$ can be defined as

$$
r(t) = E(X-t\mid X\ge t)=\frac{\int_t^\infty(u-t)f(u)du}{S(t)}
$$

where $f(t)$ is the density function.

We know that the continuous survival function with a finite mean is uniquely determined by the mean residual time as

$$
S(t)=\frac{r(0)}{r(t)}\exp \Big(-\int_0^t \frac{1}{r(u)}du \Big)
$$

Thus, as the hazard function $\lambda(t)$ can be determined by the survival function

$$
\begin{aligned}
\lambda(t)=\frac{f(t)}{S(t)}=\frac{-\frac{d}{dt}S(t)}{S(t)}=-\frac{d}{dt}\log(S(t))
\end{aligned}
$$

Where $\frac{d}{dt}S(t) = \frac{d}{dt}1 - F(t) = -\frac{d}{dt}F(t) = -f(t)$.

$\lambda(t)$ is uniquely determined by $r(t)$ by substituting $S(t)$ as

$$
\begin{aligned}
-\frac{d}{dt}\log(S(t))&=-\frac{d}{dt}\log\Bigg(\frac{r(0)}{r(t)}\exp \Big(-\int_0^t \frac{1}{r(u)}du \Big)\Bigg)\\
&=-\frac{d}{dt}\Big(\log(r(0))-\log(r(t))-\int_0^t \frac{1}{r(u)}du \Big)\\
&=-\frac{d}{dt}\Big(\log(r(0))\Big)+\frac{d}{dt}\Big(\log(r(t))\Big)+\frac{d}{dt}\Big(\int_0^t \frac{1}{r(u)}du \Big)\\
&=-(0)+\frac{\frac{d}{dt}r(t)}{r(t)}+\frac{1}{r(t)}dt\\
&=\frac{\frac{d}{dt}r(t)+1}{r(t)}\\
\end{aligned}
$$

#### b)

*Further, show that the exponential distribution is the only continuous distribution for which the mean residual lifetime $r(t)$ is constant for all $t > 0$*.

We know that the mean residual time $r(t)$ can also be expressed as

$$
r(t) = E(X-t\mid X\ge t)=\frac{\int_t^\infty S(u)du}{S(t)}
$$

Thus, with the exponential survival function for $t\ge0$ we get

$$
\begin{aligned}
r(t) = \frac{\int_t^\infty S(u)du}{S(t)} &= \frac{\int_t^\infty \exp(-\lambda u)du}{\exp(-\lambda t)} \\
&= \frac{\frac{1}{\lambda} \exp(-\lambda t)}{\exp(-\lambda t)} \\
&= \frac{1}{\lambda} \\
\end{aligned}
$$

Which of course is the mean of the exponential distribution $E(T)$. This means that the residual life $(X-t) \mid X > t$ has the same distribution as $X$. This is because the memorylessness property of the exponential distribution which states that for all $x, t \ge 0$

$$
P(X > t + x \mid X > x) = P(X > t)
$$

In our case 

$$
\begin{aligned}
P(X > t + x \mid X > x) &= \frac{S(t+x)}{S(x)} \\
&=\frac{\exp(-\lambda(t+x))}{\exp(-\lambda x)}\\
&= \exp(-\lambda t) = P(X > t)
\end{aligned}
$$

#### Q2

#### a)

**Define at-risk process and counting process for time-to-failure data.**

Let $X_1,\dots,X_n$ be a sample of uncensored continuously distributed survival times. Denote survival function at time $t$ as $S(t)=P(X > t)$ and hazard as as $\alpha(t)dt=P(t\le X < t+dt \mid X\ge t)$. Thus, naturally the cumulative hazard is $A(t)=\int_0^t \alpha(s)ds$. As we know the hazard function defines the survival function by $\exp(-\int_0^t \alpha(s)ds)$.

The data itself can be defined as

$$
(\tilde{X}_i, D_i)=\Bigg\{
\begin{aligned}
X_i=\tilde{X}_i, \quad D_i&=1 \\
X_i>\tilde{X}_i, \quad D_i&=0 \\
\end{aligned}
$$
for $i=1,\dots,n$, where $D_i$ denotes censoring for individual $i$. When $D_i=1$, $\tilde{X}_i$ is the survival time and when $D_i=0$, $\tilde{X}_i$ is the censoring time.

Let $\mathcal{F}_t$ denote filtration, as in everything that is known at time $t$. This means that $\mathcal{F}_t$ is a set of realisations of random variables prior to and including $t$. $\mathcal{F}_t-$ denotes what is known just prior to but not including $t$. I guess $\mathcal{F}_t-$ could be thought of as a limit $\lim\limits_{\epsilon\rightarrow0}\mathcal{F}_{t-\epsilon}$.

We can condition on filtration to derive, what resembles a delta function. For the hazard rate we get

$$
P(t \le \tilde{X}_i < t + dt,\, D_i = 1 \mid \mathcal{F}_t-)=\Bigg\{
\begin{aligned}
\alpha(t)dt, \quad  \tilde{X}_i &\ge t \\
0, \quad \tilde{X}_i &< t \\
\end{aligned}
$$

So the expectation for the sum over $n$ indicator functions $1\{t\le\tilde{X}_i<t+dt,\,D_i=1\}$ given $\mathcal{F}_t-$ yields what can be described the *intensity*.

$$
\begin{aligned}
E(\sum_{i=1}^n1\{t\le\tilde{X}_i<t+dt,\,D_i=1\} \mid \mathcal{F}_t-) &= \sum_{i=1}^n 1\{\tilde{X}_i\ge t\} \alpha(t)dt \\
&= \sum_{i=1}^n Y_i(t) \, \alpha(t)dt \\
&= Y(t) \, \alpha(t)dt \\
&= \lambda(t)dt \\
\end{aligned}
$$

where $Y_i(t)$ for individual $i$ is 1 if $i$ is at risk at time $t$ and is 0 otherwise, thus the intensity for individual $i$ is the hazard rate when the individual is at risk and zero when the event has happened. $Y(t)$ then denotes the size of the risk set at time $t$.

This defines the *at-risk process for time-to-failure data*.

We then define the *counting process for time-to-failure data*, that is, has an event happened prior to or at $t$.

As $Y_i(t)=1\{\tilde{X}_i\ge t\}$ denotes the individuals status on being at risk at time $t$, a counting process

$$
N_i(t)=1\{\tilde{X}_i\le t,\quad D_i=1\}, \quad 0 \le t \le \tau
$$

denotes whether the event has happened prior to or at $t$ or not. 

An increment of $N_i(t)$ is $dN_i(t)=N_i(t)-N_i(t-)$ and, as $N_i(t)$ is an indicator function, maps to either 1 on event or 0 otherwise. For a cohort of $n$ individuals $\sum_{i=1}^n N_i(t)=N(t)$ denotes the failure counting process.


#### b)

**What is the expectation of the increment conditional on the history of the process.**

An increment over a small interval $[t,t+dt)$ is 

$$
dN(t)=N((t+dt)-)-N(t-) =\sum_{i=1}^n1\{t\le\tilde{X}_i< t+dt,\quad D_i=1\}
$$

thus the expected value of increment given the history (filtration prior to $t$) is, as discussed earlier in this exercise,

$$
\begin{aligned}
E(dN(t) \mid \mathcal{F}_t-)&= E(\sum_{i=1}^n1\{t\le\tilde{X}_i< t+dt,\quad D_i=1\} \mid \mathcal{F}_t-)\\
&= \sum_{i=1}^n 1\{\tilde{X}_i\ge t\} \alpha(t)dt \\
&= \sum_{i=1}^n Y_i(t) \, \alpha(t)dt \\
&= Y(t) \, \alpha(t)dt \\
&= \lambda(t)dt \\
\end{aligned}
$$

#### Data

*Consider a study of 30 cases with complex pulmonary atresia collected as the basis of an observational study on the presentation and natural history of this disease.*

The variables are descrined below.

\begin{itemize}
 \item agepres: age at presentation
 \item agelast: age last seen alive (if alive) or age at death (if dead)
 \item ageopl: age at first operation (-1 for no operation to date)
 \item dead: no=0, yes=1
 \item sex: male=0, female=1
 \item paanat: size of intrapericardial pulmonary arteries at presentation: absent or tiny=0, normal or near normal=1
 \item adfol: adequate follow-up. Study closed less than 1 year since presentation=0, study closed at least 1 year since presentation=1
 \item Follow.up: duration of follow-up (agelast-agepres)
\end{itemize}

### Q3

Read the data as a data frame:
```{r}
cpa <- read.csv(file = "complex-pulmonary-atresia-data.csv", sep = ";")
```

#### a)

*How many patients completed the follow-up of one year and how many did not?*

The variable `Follow.up` represents the follow up time in days from age at presentation to age at death (event) or last seen alive (censoring).

Let's produce a table how many of the patients had a follow time of at least a year:
```{r}
table(cpa$Follow.up >= 365)
```

17 patients completed the one-year follow up and 13 patients did not.

#### b)

*What is the size of the risk set at 1 year?*

The risk set is the number of patients still being followed at 1 year, that is not yet experienced an event or censored.

```{r}
sum(cpa$Follow.up >= 365)
```

The size of the risk set is 17.


#### Q4

*To answer the question related to the survival experience of all patients from presentation, specify the following.*

\begin{itemize}
  \item inclusion criteria
    \begin{itemize}
      \item Inclusion criteria are the properties and characteristics that an individual must have to be included in a study.
    \end{itemize}
  \item outcome
    \begin{itemize}
      \item Outcome is the individuals **time to event**, as in time from entry to the occurrence of an event. 
    \end{itemize}
  \item time origin
    \begin{itemize}
      \item Time origin of a study is the time regarded as $T=0$. For example birth, time of diagnosis or the time of entry to the study. Basically any fixed date or time.
    \end{itemize}
  \item entry time
    \begin{itemize}
      \item Entry time, as hinted above, is the time of entry to the study. Individuals aren't always followed from time 0. It can then be regarded as delayed entry or left truncation.
    \end{itemize}
  \item censoring rule
    \begin{itemize}
      \item answer
    \end{itemize}
  \item survival time
    \begin{itemize}
      \item Survival time is the time an indiviudal spends at risk before an event.
    \end{itemize}
  \item period of observation
    \begin{itemize}
      \item Period of observation is the total time span that observations are gathered in a study.
    \end{itemize}
\end{itemize}

### Q5

**Estimate the survival function from presentation and plot the estimate with its confidence interval.**

We'll calculate a Kaplan-Meier estimate of the survival function. It is a non-parametric method, thus no assumptions are made about how the hazard rate varies in time. The method treats time as a succession of small *clicks*, each of which (should) contain only one event at most. Let $n$ be the initial risk set, $d_j$ the number of events at time $t_j$ and $l_j$ the number of censorings at time $t_j$.

Thus the risk set at time $t_i$ is given by

$$
y_i=n-\sum_{j=0}^{i-1}d_j-\sum_{j=0}^{i-1}l_j
$$

The conditional survival probability given $d_i$ events at time (click) $t_i$ is $p(T>t_i \mid T>t_{i-1}) = 1-\frac{d_i}{y_i}$ so the Kaplan-Meier estimator estimates the probability to survive beyond time $t$  by estimating surviving a succession of the previous event times (clicks). 

$$
\hat{S}(t)=P(T>t)=\prod_{i:t_i\le t} \Big(1-\frac{d_i}{y_i} \Big )
$$

We'll estimate the survival function using the `survfit` function from time of presentation `agepres` to time last seen alive `agelast`. `dead` yields the status for known events. We will also draw 95 % confidence intervals which, with `conf.type="plain"`, are given by 

$$
\left[\hat{S}(t) - z_{1-\alpha/2}\sqrt{Var(\hat{S}(t))},\   \hat{S}(t) + z_{1-\alpha/2}\sqrt{Var(\hat{S}(t))} \right]
$$

where $Var(\hat{S}(t)) =(\hat{S}(t))^2 \sum_{j:t_j\le t} \frac{d_j}{y_j - d_j}$ is the so called Greenwood formula. Other options are the `conf.type="log"` where the survival function is scaled by taking the log, as $\log(\hat{S}(t))$ or the complimentary log-log, as $\log(-\log(\hat{S}(t)))$ with  `conf.type="log-log"`.

As a thumb rule these are used if the confidence upper limit exceeds 1 (log-log) or the lower is less than 0 (log), but I fail to understand the precise intepretation and reprecautions of the log and log-log transformations in regards to the intepretation of the confidence limits. Hence, we use `plain`.

```{r}
cpa.km <- survfit(Surv(time = agepres, time2 = agelast, event = dead) ~ 1,
                  conf.int = 0.95,
                  conf.type = "plain",
                  type = "kaplan-meier",
                  data = cpa)
```


Below is the survival plot of the Kaplan-Meier estimate. The time is scaled to years for convenience. The tick-marks represent censoring.

```{r}
plot(x = cpa.km, mark.time = T, xscale = 365.25, xlab = "Time in years", ylab = "Survival probability", main = expression(paste("Kaplan-Meier estimate of the Survival function ", hat(S), "(t)")) )
```

The survival probability plummets in the first year and then continues to decline but significantly less rapidly. Censoring seems quite random at least by visual inspection. Note that the survival curve does not reach zero, more about that below.

#### a)

**What are the sizes of risk sets in the beginning and at the end?**

The follow up time `Follow.up` in ascending order standardizes the time scale to start from 0 for all patients (a row of time 0 must be added). The follow up is known for all patients, thus the follow up time is the time to event if event indicator `dead ` is 1 and the time to censoring otherwise. The number of patients included in the study is as well known.

We know the risk set at time $t$ to be

$$
y_i=n-\sum_{j=0}^{i-1}d_j-\sum_{j=0}^{i-1}l_j
$$

As there is one row of data per patient, the number of rows is the initial risk set size $n$. This can be verified via unique patient id's

```{r}
identical(n_distinct(unique(cpa$Patient)),
          (nrow(cpa)),
          (length(cpa$Patient))) 
```

In addition we calculate the cumulative number of events and the cumulative number of censorings at time $t$. The cumulative censorings are given by the complement of having experienced an event at time $t$. Finally add the "zero" row.

```{r}

cpa <- cpa %>%
  arrange(Follow.up) %>%
  mutate(n = n(),
         cum_dead = cumsum(dead),
         cum_censored = cumsum(1-dead),
         cum_y = n - cum_dead - cum_censored) %>%
  add_row(Follow.up = 0, n = 30, cum_dead = 0, cum_censored = 0, cum_y = n, .before = 1)

```

This provides what is basically a life table, but with cumulative values.

The first 6 rows:
```{r}
cpa %>%
  select(Patient, Follow.up, cum_y, cum_dead, cum_censored) %>%
  head()
```

The size of the risk set `cum_y` is 30 at the beginning.

```{r}
cpa %>%
  select(Patient, Follow.up, cum_y, cum_dead, cum_censored) %>%
  tail()
```

Now I am unsure about the indexing, as the *time bands* are aggregated $[a_{i-1},a_i)$ and we added the extra "zero row". In that perspective, as the intervals are closed on the left and open on the right, the last row with `Follow.up = 6402` is not taken in to account. The risk set would then be 1 in the end. On the other hand, we know that the last row is censored, thus accounting the information leaves us with risk set of 0.

#### b)

**Does the estimated survival function reach zero? If not, then why not?**

Let's inspect the estimated values of the survival function

```{r}
summary(cpa.km$surv)
```

The minimum value of the survival function is $0.1827$, thus it does not reach zero.

This is because the last three patients accounted for in the estimation did not experience an event and were censored. Censoring is generally regarded as *non-informative* so it does not affect the survival probability.

```{r}
data.frame(cpa.km$time, cpa.km$n.event, cpa.km$n.censor, cpa.km$surv) %>% tail()
```

As seen from the table, after the last observed event, the estimated survival probability remains unchanged.

#### c)

**What is the median survival time and the corresponding confidence interval?**

Let's use the `print` method to see the generic output of the `survfit` object

```{r}
print(cpa.km)
```

The median survival time (in days) is $1849$ with the 95 % confidence interval of $[123,5354]$ and in years

```{r}
v <- c(1849, 123, 5354, 1849/365.25, 123/365.25, 5354/365.25)
names(v) <- c("Median", "Lower", "Upper","Median (years)", "Lower (years)", "Upper (years)")
v
```

By a visual inspection this seems to be in concordance with the survival plot above.

#### Q6

**Would you recommend using exponential or Weibull distribution for the overall survival? Support your suggestion by using likelihood-ratio test.**

FIT EXPONENTIAL AND WEIBULL MODELS

#### Q7

**Perform a graphical check of proportionality for paanat. Interpret the results.**
