---
title: "Survival analysis week 2"
author: "Petteri Mäntymaa"
date: "March 20, 2019"
output:
    pdf_document:
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse);library(survival); library(eha);library(ggplot2);library(tinytex);
data(veteran)
```

## Exercise 2: Paramteric survival models and model checking

### 1. Properties of exponential distribution
#### 1.1. 

Let $T_{1},\dots, T_{n}$ be a random sample from a distribution with survival function $S(t) = exp\{−\lambda t\}$. Show that the distribution of $T = nmin(T1,\dots, Tn)$ is exponential with failure rate $\lambda$.

Note: you may prove that this result (in limit) holds even when $S(t)$ is such that $S(t) = 1 − \lambda t + o(t)$ as $t \rightarrow 0$ where $o(t)$ means $\frac{o(t)}{t} \rightarrow 0$ as $t \rightarrow 0$.

#### 1.2.

Show that the exponential distribution is the only continuous distribution for which the mean residual lifetime $r(t)$ is constant for all $t > 0$.

#### 1.3.

Show that the nth moment of the exponential distribution with falire rate $\lambda$ is $E(T^n) = \frac{n!}{\lambda n}$.

ANSWER:

We have the density function $f(t)=\lambda e^{-\lambda t}$ where $t>0$.

As the moment generating function for the exponential distribution we have

$$
\begin{aligned}
M_T(z) = E(e^{zT}) &= \int_0^\infty e^{zt} \lambda e^{-\lambda t} dt \\
                   &= \int_0^\infty e^{(z-\lambda)t} dt \\
                   &= \frac{\lambda}{z-\lambda}\Big/_0^\infty e^{(z-\lambda)t} \\
                   &= \frac{\lambda}{z-\lambda}\lim_{t \to \infty} \big(e^{(z-\lambda)t} - e^{(z-\lambda)0} \big) \\
                   &= \frac{\lambda}{z-\lambda} (0-1) \\
                   &= \frac{\lambda}{\lambda-z}, \,\ \forall \,\ \lambda<z 
\end{aligned}
$$

With some help from my calculus book we find that we can apply the following geometric series result:

$$
\frac{1}{1-x}=\sum_{n=0}^{\infty} x^{n},\,\, \forall\,\  |z| < 1
$$

With it we can rewrite the moment generating function as

$$
M_T(z) = \frac{\lambda}{\lambda-z} = \frac{\lambda}{\lambda-\frac{z}{\lambda}}=\sum_{n=0}^{\infty} \frac{1}{\lambda^n}z^{n}
$$

We also remember, again with a great deal of help from my calculus book, that 

$$
E(T^n)=M_T^{(n)}(0)=\frac{d^{n}}{dz^{n}}M_T(z), \,\ at \,\ z=0
$$

We then have to evaluate the $n^{th}$ derivative of the moment generating function, thus we need the Maclaurin series

$$
f(x)=\sum_{n=0}^{\infty}\frac{f^{(n)}(0)}{n!}x^n
$$

...and the fact that $E(T^n)=M_T^{(n)}(0)$ to rewrite the moment generating function as

$$
M_T(z)=\sum_{n=0}^{\infty}\frac{M_T^{(n)}(0)}{n!}z^n = \sum_{n=0}^{\infty}\frac{E(T^n)}{n!}z^n=\sum_{n=0}^{\infty} \frac{1}{\lambda^n}z^{n}
$$

With a little bit of tidying the last two (by getting rid of coefficients and multiplying both sides with $n!$) we finally get 

$$
E(T^n)=\frac{n!}{\lambda^n}
$$

Now this result is by no means my own invention and I gathered bits and pieces of the solution here and there but I did not utilize any parts of the solution that I did not bite and chew for a great deal first.

I do admit that many steps eluded me and I don't think I ever could have come up with them myself.

#### 2. Fitting exponential and Weibull model to Veteran data

Load veteran data from library(survival).

#### 2.1.

Plot a histogram of the survival times corresponding to uncensored observations `(veteran$status == 1)` as done in Exercise 1.

```{r}
veteran %>%
  ggplot() +
  aes(x = time, fill = factor(status)) +
  geom_histogram(binwidth = 40, color = "black") +
  theme_minimal() +
  labs(title = "Survival time of individuals", subtitle = "Veterans' Administration Lung Cancer study") +
  xlab("Survival time (days)") +
  ylab("Frequency")
```

> Minor addition to last weeks histogram; The colour indicates censoring status, red for censored (survival time unknown) and blue for non-censored (survival time known, i.e. event occurs in the study period).

#### 2.2.

Compare the Kaplan-Meier estimate of the survival function to

(a) Exponential distribution, and

(b) Weibull distribution

Use graphical procedure and interpret the results.

Hint: You can obtain the maximum likelihood estimates of the parameters using `weibreg()` function of `eha` package. For example, the estimate of the parameter $\lambda$ in $S(t) = exp\{−\lambda t\}$ is obtained using

ANSWER:

Obtain the estimate for parameter $\lambda$, create vector `t` and estimate the exponential survival probabilities for `t` as a vector `St.exp` with $S(t) = exp\{−\lambda t\}$
```{r}
veteran.exp0 <- weibreg(formula = Surv(time, status) ~ 1, data=veteran, shape=1)
lambda0 <- exp(-veteran.exp0$coeff[1])
t <- 1:1000
St.exp <- exp(-(lambda0*t))
df.exponential <- data.frame(t, St.exp)
```

Obtain the estimates for parameters `a` and `b`, create vector `t` and estimate the weibull survival probabilities for `t` as a vector `St.w` with $S(t) = exp\{−(\frac{t}{b})^a\}$
```{r}
veteran.weibull0 <- weibreg(formula = Surv(time, status) ~ 1, data=veteran)
b <- exp(veteran.weibull0$coeff[1])
a <- exp(veteran.weibull0$coeff[2])
St.w <- exp(-(t/b)^a)
df.weibull <- data.frame(t, St.w)
```

Lastly create the Kaplan-Meier survival curves with `survfit`: 
```{r}
veteran.kaplan <- survfit(Surv(time, status) ~ 1, conf.type = "none", type = "kaplan-meier", data = veteran)
```

```{r}
{
plot(veteran.kaplan, xlab = "t", ylab = "Survival probability")
lines(df.exponential$t, df.exponential$St.exp, lty = 2)
lines(df.weibull$t, df.weibull$St.w, lty = 3)
legend(x = "top", legend = c("Kaplan-Meier", "Exponential", "Weibull"), lty = c(1,2,3),bty = "n")
}
```

The exponential distribution seems to overestimate the survival probabilities approximately up to `t = 200` while the Weibull distribution seems to offer a more decent fit. Though the graphical inspection hardly offers any rigorous take on the matter, it is a important step in the process.


#### Model choice
#### 2.3.

Compare the above two models with the likelihood ratio test. Interpret the result.

Hint: You can extract the log-likelihood values from the output objects of function weibreg. Use the pchisq function to calculate the p-value (tail probability).

Alternative: You can calculate the likelihood ratio by using the anova command on the output objects from the two regression models using survreg.

Obtain the log-likelihood values of the models and calculate the likelihood ratio test score. The form of the test is $LRT = -2log_e(\frac{L_1(\hat{\theta})}{L_2(\hat{\theta})})$, but our likelihoods are already log-likelihoods so we will use $-2(log_e(L_1(\hat{\theta}))-log_e(L_2(\hat{\theta})))$.

```{r}
loglik.exp <- veteran.exp0$loglik[1]
loglik.w <- veteran.weibull0$loglik[1]

LRT <- -2*(loglik.exp - loglik.w)
LRT
```
 with $2 - 1 = 1$ degrees of freedom.

```{r}
pchisq(LRT, df = 1, lower.tail = F)
```

The P-value indicates that there is a (quite) significant difference between the models. 

#### 3. Simulation
#### 3.1.

Generate 100 random numbers from exponential distribution with mean 0.01 and store it in T. Before you do this, look at the help(rexp) and the parameter which it accepts.

```{r}
options("scipen"=0, "digits"=7)
T <- rexp(100,100)
mean(T)
```

#### 3.1.1.

Plot the empirical distribution function.

```{r}

```

#### 3.1.2.

Esimate the rate (stored under obsrate) from the simulated data and overlay the plot of the distribution function 1 − exp(−obsrate ∗ t).

#### 3.1.3.

Overlay the plot of the true exponential distribution function.

#### 3.1.4.

Explore the possibilities for different kinds of line and point plots. Vary the plot symbol, line type, line width, and colour. Also, try to give legend in the above graph.


