---
title: "Survival analysis: Week 4"
author: "Petteri Mäntymaa"
date: "May 20, 2019"
header-includes:
   - \usepackage{mathspec}
   - \usepackage{tikz}
output:
    pdf_document:
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse");library(ggplot2);library(survival);library(broom);library(eha); library(tinytex); library(survminer);library(KMsurv); library(timereg); library(boot)
options("scipen"=0, "digits"=7)
set.seed(42)
```

# Proportional hazards models and model checking

With proportional hazards models we are interested in the effect of a covariate as a scalar in regard to the hazard rate. The model consists of a failure time $T_0$ for a control group 0 with a *baseline* hazard rate of $\lambda_0(t)$ and a failure time $T_1$ for a treatment group 1 with hazard rate $\lambda_1(t)$. The hazard ratio is then

$$
\frac{\lambda_1(t)}{\lambda_0(t)} = \theta
$$
for all $t$.

According to the proportional hazards assumption, $\theta$ is constant for all $t$. Let $Z$ be a binary covariate. The proportional hazards model can be written as

$$
\lambda(t;\theta,Z) = \theta^Z\lambda_0(t)
$$

To satisfy $\theta \ge 0$, let $\beta=\exp(\theta)$. Thus

$$
\begin{aligned}
\lambda(t;\theta,Z) &= \exp(\beta Z)\lambda_0(t) \\
\end{aligned}
$$

and the hazard ratio would then be

$$
\frac{\lambda_1(t;\theta,Z_1)}{\lambda_0(t;\theta,Z_0)}=\frac{\exp(\beta Z_1)\lambda_0(t)}{\exp(\beta Z_0)\lambda_0(t)}= \exp(\beta(Z_1-Z_0))
$$


#### 3.

**Read in `melanoma` data from package `boot`. Please see the description of the data and summary of the data.**

```{r}
data(melanoma)
#?melanoma
```

The R help file consists of detailed description of the data. In short the data consists of measurements on patients with malignant melanoma taken during or after operations including follow-up information.

The data contains the following variables:

* `time`: Survival time in days since the operation, possibly censored.
* `status`: 1 = died from melanoma, 2 = still alive, 3 = death unrelated to melanoma
* `sex`: 1 = male, 0 = female
* `age`: Age in years at the time of the operation
* `year`: Year of operation
* `thickness`: Tumour thickness in mm
* `ulcer`: 1=present, 0=absent

```{r}
summary(melanoma)
```

#### 4.

**Consider ulcer as the only covariate and create Nelson-Aalen plots for patients with and without ulcer. [Hint: use `coxph()` and `survfit()`]**

We'll first create a distinct indicator variable for dying from melanoma

```{r}
melanoma$event <- ifelse(melanoma$status == 1, 1, 0)
```

and then fit a proportional hazards model stratified to baseline hazard when `ulcer==0` and hazard when `ulcer==1`. We'll also create a `survfit` object for plotting the Nelson-Aalen plots.

```{r}
mel.cox <- coxph(Surv(time, event) ~ strata(ulcer), data = melanoma)
mel.surv <- survfit(mel.cox)
```

Now for the Nelson-Aalen plot, where $\hat{\Lambda}(t)=\sum_{t_i\le t}\frac{d_i}{y_i}$ is the Nelson-Aalen estimator with with $d_i$ the number of events and $y_i$ the number individuals at risk at $t_i$

```{r}
{
plot(mel.surv, fun="cumhaz", xlab = "t", ylab = expression(Lambda(t)), col = c("black", "red"), lty = c(1,1))
legend("right",c("Ulcer == 0","Ulcer == 1"), col = c("black", "red"), lty = c(1,1))
}
```

The curves give an estimate of the cumulative hazards' shapes. There is much higher mortality with the patients with ulcerated tumors.

#### 5.

**Fit then a Cox model with ulcer as the only covariate and plot the model based estmates of the cumulative hazards in the same plot.**

```{r}
mel.cox2 = coxph(Surv(time, event) ~ ulcer, data = melanoma)
ulcer_df <- expand.grid(ulcer = c(0,1))
mel.surv2 = survfit(mel.cox2, newdata = ulcer_df)
{
plot(mel.surv, fun="cumhaz", xlab = "t", ylab = expression(Lambda(t)), col = c("black", "red"), lty = c(1,1))
legend("right",c("Ulcer == 0","Ulcer == 1"), col = c("black", "red"), lty = c(1,1))
lines(mel.surv2, fun = "cumhaz", conf.int=F, col = c("black", "red"), lty=c(2,2))
}

```

We can see the dashed curves, which are plotted from the model, following a similar path than the Nelson-Aalen estimator.

#### 6.

**Consider a Cox model with ulcer and log2(`thickness`). Plot cumulative hazards for four covariate combinations (`ulcer = 0`, `thickness = 1`, `ulcer = 0`, `thickness = 4`, `ulcer = 1`, `thickness = 4`, `ulcer = 1`, `thickness = 10`). Interpret the results. Plot also survival functions and Interpret.**

Create a new variable $\log_2(\text{thickness})$

```{r}
melanoma$log2thickness <- log2(melanoma$thickness)
```

Fit a Cox model with ulcer and $\log_2(\text{thickness})$ as covariates

```{r}
mel.cox3 <- coxph(Surv(time, event) ~ ulcer + log2thickness, data = melanoma)
```

Form a data frame from all of the combinations of ulcer and the given values for $\log_2(\text{thickness})$

```{r}
ulcer_thick <- expand.grid(ulcer = c(0,1), log2thickness = c(1,4,4,10))
```

Create the `survfit` objects for the cumulative hazards 

```{r}
mel.surv3 <- survfit(mel.cox3, newdata = ulcer_thick)
```

Plot all of the stratas

```{r}
{
plot(mel.surv3, fun="cumhaz", xlab = "t",ylab=expression(Lambda(t)),lty=1:8, col = rev(2*1:8))
legend("right",c("(0,1)","(1,1)","(0,4)","(1,4)","(0,4)","(1,4)","(0,10)","(1,10)"), col = rev(2*1:8), lty=1:8)
}
```

I ended up plotting all the combinations, but as seen from the plot on smaller log-thicknesses the differences are trivial. Thickness has a big impact on the cumulative hazard and it is multiplied by ulcer. I am aware that the y-axis scale is off. I couldn't figure out how to scale it properly.

We can easily plot the survival probabilities by changing the `fun` argument to "surv"

```{r}
{
plot(mel.surv3, fun="surv", xlab = "t",ylab=expression(hat(S)(t)),lty=1:8, col = rev(2*1:8))
legend("right",c("(0,1)","(1,1)","(0,4)","(1,4)","(0,4)","(1,4)","(0,10)","(1,10)"), col = rev(2*1:8), lty=1:8)
}
```

#### 7.

**Fit a model with ulcer as a stratify variable and log2-thickness as a covariate. Plot the cumuative baseline hazards for the two ulcer strata.**

#### 8.

**Perform graphical check of proportionality for each one of them in the model with both as covariates. Interpret the results.**

#### 9.

**Perform a formal test for proportionality by adding time-dependent covariate x log(t) and testing whether the coefficients of the time dependent covariates are significant using likelihood ratio test. How do the results compare with the graphical test results? [Hint: coxph() and cox.zph()]**
