---
title: "Untitled"
author: "Petteri Mäntymaa"
date: "March 27, 2019"
output:
    pdf_document:
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse");library(ggplot2);library(survival);library(broom);library(eha); library(tinytex)
options("scipen"=0, "digits"=7)
```


### Weibull distribution

#### 1.

How is the empirical check for the Weibull distribution performed in terms of its survival function?

ANSWER:

The Weibull model hosts one more parameter (shape) than the exponential that accounts for the increase or decrease of the hazard rates.

The survival function (under the parametrization of the course material) of the Weibull model is

$$
S(t)=\exp(-(\alpha t)^p)
$$

where $\alpha$ is the rate parameter and $p$ is the shape parameter.

By taking a log-log transformation of the Weibull survival function we get

$$
\log(-\log(S(t)))=p \log(\alpha)+p \log(t)
$$

This is linear over log(time) so based on the Kaplan-Meier estimate of the Weibull survival function, it can be applied for empirical inspection and model checking.

#### 2.

Use the result of (1) for checking the Weibull assumption for the failure time in Veteran data (Use Kaplan-Meier estimate obtained in Exercise 1)..

ANSWER:
-


#### 3.

Further

(a) repeat the Weibull assumpation check separately for the treatment groups.

(b)
Also, check the hypothesis whether the data from the treatment groups can be assumed to be drawn from the same Weibull distribution or not [use likelihood ratio test].

ANSWER:

-

#### 4.

Show that the Weibull distribution can be expressed as $Y = \alpha + \sigma W$ where $Y = log(T)$, $\alpha = − log(\lambda)$ and $\sigma = \gamma^{−1}$ and $T$ has Weibull distribution with survival function $S(t) = exp\{−(\lambda t)^\gamma\}$.

ANSWER:



#### 5.

Perform log-rank test to compare

(1) treatments

(2) prior treatment

(3) cell types.

### Fitting exponential and Weibull regression models to Veteran data

Load veteran data from library(survival).

```{r}
data(veteran)
```

#### 1.

Analyse the data with an exponential regression model using treatment status, performance status, prior therapy status and cell type as the explanatory variables.


ANSWER:

A proportional hazards exponential regression for individual $i$ is defined by the hazard rate

$$
\lambda(t_i, Z_i, \theta)=\lambda_0 (t;\theta)\exp(\beta'Z_i) = \lambda \exp(\beta'Z_i)
$$

where $\beta'$ is the transpose of the coefficient vector $\{\beta_1,\dots,\beta_k\}$ and $Z_i$ the the covariate vector $(Z_{i1},\dots,Z_{ik}),$ for individual $i$ both with $k$ the number of covariates. Also noting the constant baseline hazard $\lambda_0 (t;\theta)=\lambda$.

In our case the number of covariates would seem to be $k = 4$, though I actually assume the categorical one(s) with more than two categories will be coded as dummy variable(s), each category as a separate binary variable.

* `trt`: Treatment group, 1=standard 2=test
* `karno`: Karnofsky performance score, 100=good
* `prior`: Prior therapy, 0=no, 10=yes
* `celltype`: Cell type, 1=squamous, 2=smallcell, 3=adeno, 4=large

```{r}
exp.fit <- survreg(Surv(time, status) ~ trt + karno + prior + celltype, veteran, dist = "exponential")
summary(exp.fit)
```

#### 2.

Repeat the above analysis using Weibull model. [Use the weibreg() function].

ANSWER:
A proportional hazards Weibull regression for individual $i$ is defined by the hazard rate

$$
\lambda(t_i;Z_i,\theta)=p \alpha^p t^{p−1} \exp(\beta'Z_i)
$$

```{r}
wb.fit <- weibreg(Surv(time, status) ~ trt + karno + prior + celltype, veteran)
summary(wb.fit)
```

#### 3.

Interpret the results for the regression parameters. Which category of cell type is used as a reference category?

ANSWER:

Conserning the exponential regression model, we must note that according to the documentation (and the fact that we have negative coefficients) `survreg` reports the $\hat{\beta}$ coefficients as log-parameters so what we want is $\exp(\hat{\beta})$

```{r}
exp(exp.fit$coefficients)
```

If my understanding is correct, for example `karno` $\approx 1.03$ is the ratio of the hazards between individuals whose values of `karno` differ by one unit while other covariates are held constant. Binary variables would then be interpreted with a change from zero to one.

I am uncertain how the `celltype` is interpreted. On both models the dummy level `squamous` is left as the reference category. I would make sense for all of the three other categories to be *referenced* against the level squamous.

The both models are expressing very similar results both in the significance of the covariates (Wald test p-values) and overall p-value (goodness-of-fit: $\chi^2$-test) which is very close to zero on both.

The similarity is explained by the fact that the Weibull regression shape parameter estimate is 1.075 (as we have learned the exponential is a special case of Weibull with shape = 1)

