---
title: "Untitled"
author: "Petteri Mäntymaa"
date: "March 27, 2019"
header-includes:
   - \usepackage{tikz}
output:
    pdf_document:
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse");library(ggplot2);library(survival);library(broom);library(eha); library(tinytex)
options("scipen"=0, "digits"=7)
```


## Exercise 5: Competing risks as a multi-state model

Competing risks simulation and analysis

### 1.

Consider a competing risk model with state ”alive” and two absorbing states. See e.g. *Figure 2* in Keiding (Discussion article 1).

Simulate:

$N = 1000$ control individuals with constant cause specific hazards $\alpha_1 = 0.2$ and $\alpha_2 = 0.4$ 
$N = 1000$ treated individuals with cause-specific hazards $1.5\alpha_1$ and $1.5\alpha_2$

Censor all observations at time $T = 3$ (end of follow-up).

**Answer**

The survival model has entry state 0 and two absorbing states:
\begin{tikzpicture}[transform shape, node distance=2cm,
squarednode/.style={rectangle, draw=black, very thick, minimum size=5mm},
arrow/.style = {semithick}]
\node[squarednode](a)[align=center]{0};
\node[squarednode](b)[right of =a, align=center]{1};
\node[squarednode](c)[below of =b, align=center]{2};
\draw [->] (a) -- node [above] {$\alpha_1(t)$} (b);
\draw [->] (a) -- node [above] {$\alpha_2(t)$} (c);
\end{tikzpicture}

First let's use `rexp` to simulate draws from the exponential distribution $f(x) = \lambda {e}^{-\lambda x}$ with the given rates as in $\alpha=\frac{1}{\lambda}$. Statuses are set according to which failure occurred first. Survival times above 3 were assigned with according status.

Set parameters:
```{r}
set.seed(42)
n <- 1000
alpha1 <- 0.2
alpha2 <- 0.4
theta <- 1.5
```

Simulate `control` and `treatment` groups separately, both with `t1`'s with rate $\alpha_1$ and `t2`'s with rate $\alpha_2$, and set the status indicators accordingly (0,1 or 2)

```{r}
cont_t1 <- rexp(n = n, rate = alpha1)
cont_t2 <- rexp(n = n, rate = alpha2)
censor_t <- 3

cont_t_min <- pmin(cont_t1, cont_t2, censor_t)
cont_status <- ifelse(cont_t_min==cont_t1, 0, 1)
cont_status <- ifelse(cont_t_min >= censor_t, 2, cont_status)

treat_t1 <- rexp(n = n, rate = theta*alpha1)
treat_t2 <- rexp(n = n, rate = theta*alpha2)
treat_t_min <- pmin(treat_t1, treat_t2, censor_t)

treat_status <- ifelse(treat_t_min==treat_t1, 0, 1)
treat_status <- ifelse(treat_t_min >= censor_t, 2, treat_status)

```

### 2.

Record the simulated data in a data matrix, in which each individual’s data constitute one row with entries 0 (entry time; in this example always 0), T exit (exit time), and D (event mark with values 0, 1, or 2 for censoring and the two transitions, respectively).

**Answer**

Create data frames for `control` and `treatment` with the first column `Entry` as 0 for all observations, assign column names and bind the data frames together.

```{r}
df_cont <- data.frame(rep(0,n), cont_t_min, cont_status)
df_cont$group <- 0
names(df_cont) <- c("Entry", "T", "D", "Group")

df_treat <- data.frame(rep(0,n), treat_t_min, treat_status)
df_treat$group <- 1
names(df_treat) <- c("Entry", "T", "D", "Group")

df <- rbind(df_cont, df_treat)

```

N.B. The true model employs an assumption of constant baseline hazards in the two strata (control and treated individuals), and a shared multiplicative effect of treatment.

#### a.

Prepare your data for a competing risk analysis where you assume that both transitions have their own hazards and the effect of treatment is different on different transitions.

Fit an exponential survival regression model to the simulated data.

**Answer**

Separate the data frames so that data frame 1 has status indicator with 1 for status 1 and 0 for the rest and data frame 2 has status indicator with 1 for status 2 and 0 for the rest. As discussed in the lecture material in each data frame holds the status only for cause-specific hazard in question and failure by any of the other cause is taken into account as censoring.

```{r}
df1 <- df
df2 <- df

df1$D <- ifelse(df1$D == 1,1,0)
df2$D <- ifelse(df2$D == 2,1,0)
```

Thus we have 1158 events in data 1 and 243 events in data 2.
```{r}
sum(df1$D)
sum(df2$D)
```

Fit an exponential model with `weibreg` by assigning the shape argument as 1. In the documentation the model is spesified as

$$
\begin{aligned}
h(t; a, b, \beta, z) &= \frac{a}{b} (\frac{t}{b})^{a-1} exp(z\beta)\\
h(t; a=1, b, \beta, z) &= \frac{1}{b} (\frac{t}{b})^{1-1} exp(z\beta)\\
h(t; b, \beta, z) &= \frac{1}{b} exp(z\beta)\\
\end{aligned}
$$

```{r}
w.fit1 = weibreg(Surv(T, D) ~ Group,data=df1,shape=1)
summary(w.fit1)
```


```{r}
w.fit2 = weibreg(Surv(T, D) ~ Group,data=df2,shape=1)
summary(w.fit2)
```


#### b.

Prepare the same data for a competing risk analysis where you now assume that both transitions have their own baseline hazards, but the (multiplicative) effect of treatment is the same on both transitions. 

Analyse the data. You can try both the weibreg and coxph functions.


**Answer**

We can use the previous data, but after assigning statuses to 0/1 combine the data again and add a variable for stratification.

```{r}
df1$strat <- 1
df2$strat <- 2
df_strat <- rbind(df1,df2)
df_strat$strat <- factor(df_strat$strat)

```

```{r}
ws.fit <- weibreg(Surv(T, D) ~ Group+strat,data=df_strat,shape=1)
summary(ws.fit)
```

#### c.

Prepare the same data for a competing risk analysis where you assume proportional baseline hazards and different effects of treatment on both transitions. Analyse the data. At each analysis, compare your results to the values of the model parameters that you used to simulate the data. Because the size of the data is considerable, the point estimates should be reasonably close to these “true” values


**Answer**

Finally, combine the both treatment groups to the same data frame and again add the stratification indicator accordingly

```{r}
dfp1 <- df
dfp2 <- df

dfp1$D <- ifelse(dfp1$D == 1,1,0)
dfp1$Group1 <- dfp1$Group 
dfp1$Group2 <- 0
dfp1$strat <- 1

dfp2$D <- ifelse(dfp2$D == 2,1,0)
dfp2$Group1 <- 0
dfp2$Group2 <- dfp2$Group
dfp2$strat <- 2

df_prop <- rbind(dfp1,dfp2)
df_prop$strat <- factor(df_prop$strat)

```

```{r}
wsp.fit <- weibreg(Surv(T, D) ~ Group1+Group2+strat,data=df_prop,shape=1)
summary(wsp.fit)
```

