---
title: "Survival week 5 REVISED"
author: "Petteri Mäntymaa"
date: "May 6, 2019"
header-includes:
   - \usepackage{tikz}
output:
    pdf_document:
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse");library(ggplot2);library(survival);library(broom);library(eha); library(tinytex)
options("scipen"=0, "digits"=7)
```


## Exercise 5: Competing risks as a multi-state model

Competing risks simulation and analysis

### 1.

Consider a competing risk model with state ”alive” and two absorbing states. See e.g. *Figure 2* in Keiding (Discussion article 1).

Simulate:

$N = 1000$ control individuals with constant cause specific hazards $\alpha_1 = 0.2$ and $\alpha_2 = 0.4$ 
$N = 1000$ treated individuals with cause-specific hazards $1.5\alpha_1$ and $1.5\alpha_2$

Censor all observations at time $T = 3$ (end of follow-up).

**Answer**

Our survival model is a multi-state model with entry state 0 and two absorbing states:

\begin{tikzpicture}[transform shape, node distance=2cm,
squarednode/.style={rectangle, draw=black, very thick, minimum size=5mm},
arrow/.style = {semithick}]
\node[squarednode](a)[align=center]{0};
\node[squarednode](b)[right of =a, align=center]{1};
\node[squarednode](c)[below of =b, align=center]{2};
\draw [->] (a) -- node [above] {$\alpha_1(t)$} (b);
\draw [->] (a) -- node [above] {$\alpha_2(t)$} (c);
\end{tikzpicture}

First let's use `rexp` to simulate draws from the exponential distribution $f(x) = \lambda {e}^{-\lambda x}$ with the given rates as in $\alpha=\frac{1}{\lambda}$. Statuses are set according to which failure occurred first. Survival times above 3 were assigned with according status.

Set parameters:
```{r}
set.seed(42)
n <- 1000
alpha1 <- 0.2
alpha2 <- 0.4
trtcoef <- 1.5
```

Simulate `control` and `treatment` groups separately, both with `t1`'s with rate $\alpha_1$ and `t2`'s with rate $\alpha_2$, and set the status indicators accordingly (0,1 or 2)

```{r}
cont_t1 <- rexp(n = n, rate = alpha1)
cont_t2 <- rexp(n = n, rate = alpha2)
censor_t <- 3

cont_t_min <- pmin(cont_t1, cont_t2, censor_t)
cont_status <- ifelse(cont_t_min==cont_t1, 0, 1)
cont_status <- ifelse(cont_t_min >= censor_t, 2, cont_status)

treat_t1 <- rexp(n = n, rate = trtcoef*alpha1)
treat_t2 <- rexp(n = n, rate = trtcoef*alpha2)
treat_t_min <- pmin(treat_t1, treat_t2, censor_t)

treat_status <- ifelse(treat_t_min==treat_t1, 0, 1)
treat_status <- ifelse(treat_t_min >= censor_t, 2, treat_status)

```

### 2.

Record the simulated data in a data matrix, in which each individual’s data constitute one row with entries 0 (entry time; in this example always 0), T exit (exit time), and D (event mark with values 0, 1, or 2 for censoring and the two transitions, respectively).

N.B. The true model employs an assumption of constant baseline hazards in the two strata (control and treated individuals), and a shared multiplicative effect of treatment.

**Answer**

Create data frames for `control` and `treatment` with the first column `Entry` as 0 for all observations, assign column names and bind the data frames together.

```{r}
df_cont <- data.frame(rep(0,n), cont_t_min, cont_status)
df_cont$group <- 0
names(df_cont) <- c("Entry", "T", "D", "Group")

df_treat <- data.frame(rep(0,n), treat_t_min, treat_status)
df_treat$group <- 1
names(df_treat) <- c("Entry", "T", "D", "Group")

df <- rbind(df_cont, df_treat)

```

#### a.

Prepare your data for a competing risk analysis where you assume that both transitions have their own hazards and the effect of treatment is different on different transitions.

Fit an exponential survival regression model to the simulated data.

**Answer**

Separate the data frames so that data frame 1 has status indicator with 1 for status 1 and 0 for the rest and data frame 2 has status indicator with 1 for status 2 and 0 for the rest. As discussed in the lecture material in each data frame holds the status only for cause-specific hazard in question and failure by any of the other cause is taken into account as censoring.

```{r}
df1 <- df
df2 <- df

df1$D <- ifelse(df1$D == 1,1,0)
df2$D <- ifelse(df2$D == 2,1,0)
```

Thus we have 1158 events in data 1 and 243 events in data 2.
```{r}
sum(df1$D)
sum(df2$D)
```

We'll fit exponential models for both data with `weibreg` (which we get from Weibull distribution by assigning the shape argument as 1).


```{r}
w.fit1 = weibreg(Surv(T, D) ~ Group,data=df1,shape=1)
summary(w.fit1)
```

```{r}
w.fit2 = weibreg(Surv(T, D) ~ Group,data=df2,shape=1)
summary(w.fit2)
```

We are seeing a quite close estimate of the 1.5 treatment coefficient in Exp(Coef) for Group1 but not Group2. Possibly because the first data holds over 1000 events while the latter some 240.

#### b.

Prepare the same data for a competing risk analysis where you now assume that both transitions have their own baseline hazards, but the (multiplicative) effect of treatment is the same on both transitions. 

Analyse the data. You can try both the weibreg and coxph functions.


**Answer**

We can use the previous data, but after assigning statuses to 0/1 combine the data again and add a variable for stratification.

```{r}
df1$strat <- 1
df2$strat <- 2
df_strat <- rbind(df1,df2)
df_strat$strat <- factor(df_strat$strat)

```

The baseline hazards are now estimated by each stratum but the treatment effect is overlooked

```{r}
ws.fit <- weibreg(Surv(T, D) ~ Group+strat,data=df_strat,shape=1)
summary(ws.fit)
```




#### c.

Prepare the same data for a competing risk analysis where you assume proportional baseline hazards and different effects of treatment on both transitions. Analyse the data. At each analysis, compare your results to the values of the model parameters that you used to simulate the data. Because the size of the data is considerable, the point estimates should be reasonably close to these “true” values


**Answer**

Finally, combine the both treatment groups to the same data frame and again add the stratification indicator by status and treatment group.

```{r}
dfp1 <- df
dfp2 <- df

dfp1$D <- ifelse(dfp1$D == 1,1,0)
dfp1$Group1 <- dfp1$Group 
dfp1$Group2 <- 0
dfp1$strat <- 1

dfp2$D <- ifelse(dfp2$D == 2,1,0)
dfp2$Group1 <- 0
dfp2$Group2 <- dfp2$Group
dfp2$strat <- 2

df_prop <- rbind(dfp1,dfp2)
df_prop$strat <- factor(df_prop$strat)

```

The `weibreg` function now performes a proportional hazards model with baseline hazards from the exponential distribution (again Weibull distribution with a shape parameter of 1). This time we allow stratification with different scale estimated in each stratum.

```{r}
wsp.fit <- weibreg(Surv(T, D) ~ Group1+Group2+strat,data=df_prop,shape=1)
summary(wsp.fit)
```

Let's pay special attention to the estimated coefficients for the treatment groups

```{r}
c(exp(wsp.fit$coefficients[1]),exp(wsp.fit$coefficients[2]))
```

Oddly, probably as per some error in the data generation, only $\exp(\alpha)$ for `Group1` is closing on the coefficient of 1.5 assigned to the treatment groups. I suppose they both should exhibit the effect of the treatment.

In fact on closer inspection they are the same estimates obtained from the separate model exponential models in part **a**. I will try to figure this out.

